---
title: "Using R~t~ to explore firebreaks in Wales and Northern Ireland"
author: "Kath Sherratt, Sam Abbott, Epiforecasts and Seb Funk"
date: "03/11/2020"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Rationale

We aimed to explore ongoing intensive lockdowns (firebreaks) in Northern Ireland (from 16th October 2020) and Wales (24th October 2020), using the reproduction number over time (R~t~). In contrast to using a smooth Gaussian Process method, we estimated R~t~ using breakpoints.

## Methods

We used government data (shown as a 7 day moving average in Fig 1A), and used two methods to estimate R~t~ over a ten week period, and repeated each method for each of data reported as cases, hospital admissions, and deaths.

First we modelled a single breakpoint in R~t~ at the start of respective lockdowns (figure 2B). Second, we allowed for a weekly random walk in R~t~ by including a breakpoint on each Sunday, with an additional breakpoint at the start of respective lockdowns. In the latter method, we left at least one week between the last breakpoint and the firebreak breakpoint. No further breakpoints were included beyond the start of the firebreak.

We repeated this process for the South West region of England. This was included as a comparison given this region saw the lowest level of policy interventions. We included breakpoints as for Wales (i.e. 24th October).

## Results

We saw some evidence of an impact of the firebreak when R~t~ was modelled using breakpoint. The firebreak appeared to have a strong effect on decreasing R~t~ when estimated from reported cases in Northern Ireland, using either a single breakpoint or a weekly random walk with a breakpoint at the firebreak. However, the strength of this firebreak effect on R~t~ was less clear across other regions (Wales and South West) and data sources (admissions and deaths), although we observed a general decline in R~t~.

```{r, fig.height = 2, fig.width = 5, fig.align='center', echo=FALSE}
plot(readRDS(here::here("rt-estimate/estimate-break", "break-effect-size.rds")))
```

*Figure 1. Effect size of breakpoint on R~t~, on the date of the firebreak (24th October for Wales and South West, 16th October for N Ireland), median with 90% credible intervals.*

```{r, fig.height = 15, fig.width = 15, echo=FALSE}
suppressWarnings(plot(readRDS(here::here("rt-estimate/estimate-break", "sw-wales-ni.rds"))))
```

*Figure 2. A: Data for cases, hospital admissions, and deaths, in the South West, Wales, and Northern Ireland. B: R~t~ modelled using a single breakpoint: 24th October in the South West (no intervention); 24th October in Wales (firebreak); 16th October in Northern Ireland (firebreak). C: R~t~ including a random walk, with breakpoints each Sunday until the week before the firebreak, plus the single breakpoint as in (B).*
